|

### 2023-05-02 paper.Rmd

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, can be used to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. In general, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. However, correctly judging whether NO pattern exists in a residual plot is a difficult task for humans. Humans will almost always will see a pattern (see @kahneman), so the question that really needs answering is whether any pattern perceived is consistent with randomness, purely sampling variability, or noise. It is especially difficult to teach this to new analysts and students [@loy2021bringing]. To address this, statistical tests are available to check for non-linear patterns [e.g. @ramsey1969tests], and heteroskedasticity  [e.g. @breusch1979simple].

@@

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (A good discussion can be found in @kirk1996.) While such rejections may be statistically correct, their sensitivity may render the results impractical. A key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses, but minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. The experiment discussed in this paper specifically addresses this tension between statistical and practical statistics.

>>>>

In addition, scatterplots of residuals plotted against the fitted values and each of the explanatory variables, can be used to scrutinize their relationships. If there are any visually discoverable patterns, the model is potentially inadequate or incorrectly specified. In general, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. However, correctly judging whether NO pattern exists in a residual plot is a difficult task for humans. Humans will almost always will see a pattern (see @kahneman2011thinking), so the question that really needs answering is whether any pattern perceived is consistent with randomness, purely sampling variability, or noise. It is especially difficult to teach this to new analysts and students [@loy2021bringing]. To address this, statistical tests are available to check for non-linear patterns [e.g. @ramsey1969tests], and heteroskedasticity  [e.g. @breusch1979simple].

@@

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (A good discussion can be found in @kirk1996practical.) While such rejections may be statistically correct, their sensitivity may render the results impractical. A key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses, but minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. The experiment discussed in this paper specifically addresses this tension between statistical and practical statistics.
